A Simple Type Theory for Category Theory

### 2022-04-26 paper/paper.tex
Intro/abstract rewrite

\usepackage{tikz-cd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage{bbm}
\usepackage{cleveref}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to a form of ordered
  logic, we guarantee that all mappings between categories are
  functorial and all transformations are natural by construction, with
  no separate proof necessary. Important category theoretic proofs
  such as the Yoneda lemma and Co-yoneda lemma become simple type
  theoretic proofs about the relationship between unit, tensor and
  (ordered) function types.

  We show that the type theory is sound for models in \emph{virtual
  equipments}, a kind of double category which is known to model not
  just ordinary categories but enriched, internal and generalized
  multi-categories as well. This means that while proofs and
  constructions in our theory read like typical category-theoretic
  constructions, the proofs automatically apply to these generalized
  categories. Our type theory simplifies arguments over the algebraic
  description of virtual equipments by modeling the algebraic
  operation of restriction of profunctors along functors as instead
  the strict syntactic substitution.
\end{abstract}

@@

Category theory is a branch of mathematics that formalizes the
algebraic structure of function composition. The high level of
abstraction allows for succinct proofs relevant to structures across a
broad variety of mathematical domains. Category theory is of
particular interest in the theory of programming languages and proof
assistants, where it is commonly used to aid in the design of typed
languages, proof assistants and functional programming abstractions.

Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of
the Yoneda lemma, often the first ``big'' proof in an introductory
category theory text. ... Here the reader is left to themselves to
show that the ``evaluation at identity'' morphism and its inverse are
actually natural transformations. They will then work through a
diagram chase to determine that this naturality condition indeed
holds.

>>>>

\usepackage{tikz-cd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage{bbm}
\usepackage{cleveref}
\usepackage{natbib}

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to a form of ordered
  logic, we guarantee that all mappings between categories are
  functorial and all transformations are natural by construction, with
  no separate proof necessary. Important category theoretic proofs
  such as the Yoneda lemma and Co-yoneda lemma become simple type
  theoretic proofs about the relationship between unit, tensor and
  (ordered) function types. We demonstrate that this type theory can
  be used as a kind of logical framework in which we can define common
  categorical structures such as categories with families that serve
  as models of type theories.

  Finally, we give a sound and complete notion of model of our type
  theory in a structure we call a hyperdoctrine of pro-arrow
  equipments with displays, a generalization of prior notions of
  virtual equipments that have been shown to model enriched and
  internal category theory. This means that while the proofs in our
  type theory look like standard set-based arguments, the
  substructural syntactic restrictions ensure that the proofs carry
  over to enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that gives a systematic
theory of typed, ``higher-dimensional'' algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that these algebraic structures were common in
logic and computer science, and that formal systems like type theories
and programming languages can be modeled by a sound and complete
notion of semantic model given by some notion of structured
category\cite{lawvere,goguen,lambek,moggi}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{not-sure}.
%
These seminal works have led to a fruitful field of research applying
such categorical logic techniques to the design of new type theories
based directly on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in ``pure'' mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual typing.

% TODO: Yoneda example
Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of
the Yoneda lemma, often the first ``big'' proof in an introductory
category theory text. ... Here the reader is left to themselves to
show that the ``evaluation at identity'' morphism and its inverse are
actually natural transformations. They will then work through a
diagram chase to determine that this naturality condition indeed
holds.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When a
physicist writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natual, etc.
%
In the case of category theory, these principles can be formalized
using category-like algebraic structures such as 2-categories,
bicategories, (virtual) double categories and pro-arrow
equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
The downside is that these algebraic structures are quite complicated,
with a 2-dimensional diagrammatic language that can be quite beautiful
and elegant, but is also seems very far removed from the traditional
formulation of category theory that while complex is at least
described in terms of a concrete notion of sets or types.

In this work we show that these algebraic formulations of category
theory can be given a type-theoretic syntax that allows us to make
``na\"ive'' set theoretic constructions for categories, but by
adhering to a certain syntactic discipline provide correct arguments
for not just ordinary category theory but enriched and internal
categories as well.
%
The type theory, which we call SCT, is a ``synthetic'' theory of
categories, functors, profunctors and natural transformations.
%
This provides a formulation of category theory in which profunctors
play a primary role, most recognizable in ordinary category theory in
the prominence of ``hom isomorphism'' formulations of universal
properties.
%
The semantics of SCT is given by a slight generalization of virtual
double categories \cite{} that we call double categories with
displays.
%
Additionally, to define signatures and parameterize theorems like the
Yoneda lemma, SCT includes an ordinary dependent type theory over it
which includes usual $\Pi,\Sigma,\Id$ types.
%
In an implementation of SCT, this dependent type theory might be the
host language itself.

In this work, we focus on the syntax and semantics of SCT, but the
ultimate goal is for this theory to serve as the foundation for a
framework in which we can more easily study and program with new type
theories. This falls in the tradition of \emph{logical frameworks}
such as the LF system. To provide evidence for the feasibility of this
approach, we show that we can model structures used in recent work on
modal dependent type theory. This shows that an implementation of SCT
would provide at least a toy prototype implementation/type checker for
modal dependent type theory. This can be summarized in the slogan
\begin{quote}
  category theory as a logical framework
\end{quote}
that is, that a syntax for category theory can be used as a logical
framework for type theories with semantics in structured categories.
%
If implemented in a proof assistant this would also provide with it
the soundness of the syntax in any model.
%
The emphasis on profunctors in SCT fits nicely with the
\emph{judgmental} approach of Per-Martin L\"of and suggests the
following slogan
\begin{quote}
  profunctors as judgments, natural transformations as derivations
\end{quote}
because we observe that the ``derivation tree'' style of writing
judgments corresponds closely to the type structure of profunctors and
natural transformations in SCT, and it should be simple to implement a
program that converts formal natural transformations into derivation
tree syntax.

\subsection{Category theory with Profunctors}

\subsection{Double Categories and Pro-arrow Equipments}
